### 2025-09-23 paper/sec/introduction.tex
Extract figure for SciGen library.

\figref{scigen-example-website}, generated from our implementation, illustrates these two features. The upper
section shows an excerpt from a research paper comparing text encoding techniques~\citep{zhang18}. When a
reader hovers over the phrase ``does not further improve'', the corresponding data entries are highlighted in
blue on the left. Other fragments (e.g. ``better than'', ``further improvements'') that refer to the same data
are also marked, allowing readers to explore supporting and contrasting evidence. The lower section shows a
counterfactual situation where we imagine the authors' experiments had produced different results: here the
phrase ``does not further improve'' is replaced by the ``further improves''. Transparent documents thus turn
static text into interactable, data-driven content able to expose the evidential basis of scholarly claims.

@@

NLP aspect of the problem is potentially a big problem space in itself. We will restrict interest to certain
idiomatic uses of natural language in making/justifying scientific claims. Table~\ref{tab:fluid_examples}

>>>>

\figref{scigen-example-website}, generated from our implementation, illustrates these two features. The upper
section shows an excerpt from a research paper comparing text encoding techniques~\citep{zhang18}. When a
reader hovers over the phrase \textfrag{does not further improve}, the corresponding data entries are
highlighted in blue on the left. Other fragments (e.g. \textfrag{better than}, \textfrag{further
improvements}) that refer to the same data are also marked, allowing readers to explore supporting and
contrasting evidence. The lower section shows a counterfactual situation where we imagine the authors'
experiments had produced different results: here the phrase \textfrag{does not further improve} is replaced by
\textfrag{further improves}. Transparent documents thus turn static text into interactable, data-driven
content able to expose the evidential basis of scholarly claims.

@@

Table~\ref{tab:fluid_examples} summarises the natural language idioms studied in this paper. With
state-of-the-art models like gpt4o and gpt5, our system is able to resolve basic table lookups of direct
numerical values, as well as computations of percentages, averages, minima and maxima, and totals, each mapped
to the corresponding aggregation over the source data. For example, phrases such as \textfrag{the Energy
Sector accounts for 52.80\% of total emissions} or \textfrag{average methane emissions for 2030 is 13.51} are
interpreted in terms of sum and mean operations respectively over the relevant data values. Similarly,
\textfrag{recorded its highest emissions in 2030} is interpreted as a \kw{maximumBy} query, while a statement
such as \textfrag{CNN gives the lowest accuracy} is interpreted as an explicit computation of rank.

We also consider \emph{trend} expressions, which comparative natural language phrases describing how a data
attribute evolves over time, such as \textfrag{training time growing from 67 to 106 seconds}. Such idioms are
mapped to higher-order functions like \kw{trendWord} which are parameterised on additional helper functions
such as \kw{growShrink} and \kw{betterWorse} (\figref{}) which map comparisons to appropriate natural language
phrases.

\input{fig/scigen-lib}

Taken together, these categories cover a representative portion of the numerical reasoning idioms found in the
SciGen benchmark. However, some common linguistic forms are not covered in our analysis. We have yet to study
approximate quantitative terms like \textfrag{around 50\%} or \textfrag{roughly 100 instances}, nor
interval-based descriptions such as \textfrag{between 30 and 40\%} or \textfrag{within 5–10 seconds}. We have
no reason for thinking these will present specific difficulties, however. Other forms are likely to be more
challenging. So-called \emph{graded} modal adverbs~\citep{lassiter17} which modify adjectival comparatives
like ``better`` -- like \textfrag{slightly better} and \textfrag{significantly higher} -- especially when
combined with trends over time, as in \textfrag{steadily increasing} or \textfrag{sharply declining}, because
the interpretation of these qualifiers is quite subjective and contextual. Adverbial quantifiers like
\textfrag{generally} and \textfrag{usually} also present challenges because colloquial use (``most'' meaning
``greater than any other alternative proportion'' may differ from more formal uses (``most'' meaning a
majority, i.e.~greater than 50\% of cases). Extending coverage to these idioms would substantially deepen our
tool's ability to bridge natural language reporting with interpretation in terms of the underlying dataset. We
discuss this further in \secref{future-work}.
